Large and Infinitary Quotient Inductive-Inductive Types

### 2020-01-10 draft/paper.tex
paper

Algebraic notions of models of type theories are examples for this. Here,
initial models represent syntax, and initiality corresponds to induction on
syntax. Several variants have been used, from contextual categories \cite{gat}
and comprehension categories \cite{jacobs1993comprehension} to categories with
families \cite{Dybjer96internaltype} (CwF).
%
Whereas the initial category can be given by first defining the type
of objects (the empty type) and then defining morphisms as a separate
inductive family, for the initial CwF one has to define $Ob$, $Mor$
and two extra sorts $Ty$ and $Tm$ \emph{inductive-inductively}, at the
same time. This is because there is a constructor of $Ob$ which refers
to $Ty$, constructors of $Ty$ refer to $Ob$ and $Sub$, and so on. This
is illustrated by the following excerpt of the signature for CwFs
extending the above signature for categories.
\begin{alignat*}{3}
  & Ty && : Ob\ra \U \\
  & empty  && : Ob \\
  & ext && : (I:Ob)\ra Ty\,I\ra Ob \\
  & \blank[\blank] && : Ty\,J\ra Mor\,I\,J\ra Ty\,I
\end{alignat*}
While signatures for type theory usually only contain finitary constructors,
some QIITs have infinitary constructors, such as the following signature for
countable branching trees where the order of subtrees does not matter.
\begin{alignat*}{3}
  & Tree && : \U \\
  & leaf  && : Tree \\
  & node && : (\N\ra Tree)\ra Tree \\
  & perm && : (g :\N\ra Tree)(f : \N\ra\N)\ra \mathsf{isIso}\,f\ra \\
  & && \hspace{0.8em} node\,g = node (g\circ f)
\end{alignat*}
Here $node$ and $perm$ are infinitary constructurs, $perm$ expresses
that reordering subtrees by a permutation (a $\N\ra\N$ function which
is an isomorphism) does not matter.

@@

\emph{Notational conventions.} We name elements of $\Con$ as $\Gamma$, $\Delta$,
$\Theta$, elements of $\Sub\,\Gamma\,\Delta$ as $\sigma$, $\delta$, $\nu$,
elements of $\Ty\,\Gamma$ as $A$, $B$, $C$, and elements of $\Tm\,\Gamma\,A$ as
$t$, $u$, $v$. CwF components by default support de Bruijn indices, which are
not easily readable. We use instead a nameful notation for binders in context
extension, $\Pii$ and $\lami$, e.g.\ as $(\emptycon\,\ext (a : U) \ext (t :
\El\,a))$. We also define a type-theoretic flavor of $\appi$ for convenience:
\begin{alignat*}{3}
  & \blank\appitt\blank && :
      \Tm\,\Gamma\,(\Pii\,a\,B)\ra
      (u : \Tm\,\Gamma\,(\El\,a)) \ra \Tm\,\Gamma\,(B[\id, u])\\
  & t\appitt u && := (\appi\,t)[\id, u]
\end{alignat*}
We abbreviate non-dependent inductive $\Pii$ as $\blank\arri\blank$, and
likewise we use $\blank\arre\blank$ and $\blank\arrinf\blank$ for non-dependent
external and infinitary functions.

@@

With this, we are able to specify QIITs, and we can also interpret each
signature in an arbitrary ToS model, by applying a signature to a model.
$\Sig_j$ can be viewed as a precursor to a Church-encoding for the theory of
signatures, but we only need contexts encoded in this way, and not other ToS
components. In functional programming, this is sometimes called ``finally
tagless'' \cite{carette2007finally}, and it is used for defining and interpreting
embedded languages.
\end{definition}

@@

We make a similar choice in the interpretation of signatures themselves: we use
structured CwFs instead of lex categories. The reason here is that CwFs allow us
to compute induction principles in strictly the same way as one would write them
down in type theory, since we have $\Ty$ and $\Tm$ for a primitive notion of
dependent objects and morphisms. In contrast, dependent objects in lex
categories is a derived notion, and the induction principles we get are only up
to isomorphism. This issue is perhaps not relevant from a purely categorical
perspective, but we are concerned with eventually implementing QIITs in proof
assistants, so we prefer if our semantics computes strictly.

@@

In contrast to \cite{kaposi2019constructing}, it is not sufficient to model
types with displayed flCwFs. In ibid.\ substitutions are modeled as strict
morphisms, which we cannot do because our infinitary function types and identity
types cannot be interpreted strictly, as we will see in Section
\ref{sec:infinitaryfunction}. Strict preservation is expressed with metatheoretic equality, which
is respected by all cETT constructions, but isomorphisms are not necessarily
respected. Hence, we need to add additional structure to displayed flCwFs which
expresses preservation of base isomorphisms.

@@

Moreover, $\id_{\boldsymbol{A[\sigma]}} := \id_{\bA}$,
$\sigma\circ_{\boldsymbol{A[\sigma]}}\delta := \sigma\circ_{\bA}\delta$,
and likewise substitution components are given by corresponding substitution
components in $\bA$.

@@

We will need in the following three model fragments of $\bM$, which can be used
to compute notions of algebras, displayed algebras and sections respectively for
each syntactic signature. This is a rephrasing of the $\blank^{\A}$,
$\blank^{\D}$ and $\blank^{\S}$ interpretations in
\cite{kaposi2019constructing}.

@@

The basic idea is that initial algebras can be built from the terms of $\syn_j$.
For example, consider the syntactic signature for natural numbers:
\[
\NatSig := \emptycon\ext(N : \U)\ext(zero : \El\,N)\ext (suc : N
\arri \El\,N)
\]
The type $\Tm_{\syn}\,\NatSig\,(\El_{\syn}\,N)$ is isomorphic to the usual type
of natural numbers, since, intuitively, such terms can only be built from
iterated usage of $zero$ and $suc$. We will build a term algebra for each
signature in this manner. First, we describe the interpretation of $\syn$ in the
$\Set$ model.

@@

So far, the underlying sets in $\bT_{\ulOmega}$ are similar to what was given in
\cite{kaposi2019constructing} in the construction of term algebras, but there is
an important difference: in ibid.\ strict equalities are used instead of
isomorphisms. In our case, isomorphisms are necessary once again because of
infinitary functions types and our identity type; we shall see this shortly. The
universe is interpreted as follows:
\begin{alignat*}{3}
  & \U_{\bT_{\ulOmega}} : (\ulnu : \Sub\,\ulOmega\,\ulGamma)(\ult : \Tm\,\ulOmega\,\U)\ra
              \Set_{j+1}\\
  & \U_{\bT_{\ulOmega}}\,\ulnu\,\ult := \Tm\,\ulOmega\,(\El\,\ult)\\
  & \El_{\bT_{\ulOmega}}\,a : (\ulnu : \Sub\,\ulOmega\,\ulGamma)
             (\ult : \Tm\,\ulOmega\,(\El\,(\ula[\ulnu])))\ra
              \llb\ula\rrb_{\bA}\,(\Gamma\,\ulnu)\\
  & \El_{\bT_{\ulOmega}}\,a\,\ulnu\,\ult := (a\,\ulnu)\,\ult
\end{alignat*}

Hence, a syntactic $\ult : \Tm\,\ulOmega\,\U$ is interpreted as a set of terms
with type $\El\,\ult$. In the interpretation of $\El$, note that
\[
  a\,\ulnu : \U_{\bT_{\ulOmega}}\,\ulnu\,(\ula[\ulnu]) \simeq_{\id}
                 \llb\ula\rrb_{\bA}\,(\Gamma\,\ulnu)
\]
hence
\[
a\,\ulnu : \Tm\,\ulOmega\,(\El\,(\ula[\ulnu])) \simeq_{\id}
           \llb\ula\rrb_{\bA}\,(\Gamma\,\ulnu)
\]
The $\simeq_{\id}$ above is just an isomorphism of sets, since it lives in
$\llb\U\rrb_{\bM}$ which was given as the flCwF of sets in Section
\ref{sec:universe}. This above isomorphism is a good summary of the
construction: the interpretation of a $\ula : \Tm\,\ulOmega\,\U$ in the term
algebra is isomorphic to a set of terms.

>>>>

Algebraic notions of models of type theories are examples for this. Here,
initial models represent syntax, and initiality corresponds to induction on
syntax. Several variants have been used, from contextual categories \cite{gat}
and comprehension categories \cite{jacobs1993comprehension} to categories with
families \cite{Dybjer96internaltype} (CwF).
%
Whereas the initial category can be given by first defining the type
of objects (the empty type) and then defining morphisms as a separate
inductive family, for the initial CwF one has to define $Ob$, $Mor$
and two extra sorts $Ty$ and $Tm$ \emph{inductive-inductively}, at the
same time. This is because there is a constructor of $Ob$ which refers
to $Ty$, constructors of $Ty$ refer to $Ob$ and $Sub$, and so on. This
is illustrated by the following excerpt of the signature for CwFs
extending the previous signature for categories.
\begin{alignat*}{3}
  & Ty && : Ob\ra \U \\
  & empty  && : Ob \\
  & ext && : (I:Ob)\ra Ty\,I\ra Ob \\
  & \blank[\blank] && : Ty\,J\ra Mor\,I\,J\ra Ty\,I
\end{alignat*}
While signatures for type theory usually only contain finitary constructors,
some QIITs have infinitary constructors, such as the following signature for
countable branching trees where the order of subtrees does not matter.
\begin{alignat*}{3}
  & Tree && : \U \\
  & leaf  && : Tree \\
  & node && : (\N\ra Tree)\ra Tree \\
  & perm && : (g :\N\ra Tree)(f : \N\ra\N)\ra \mathsf{isIso}\,f\ra \\
  & && \hspace{0.8em} node\,g = node (g\circ f)
\end{alignat*}
Here $node$ and $perm$ are infinitary constructurs, $perm$ expresses
that reordering subtrees by a permutation (a $\N\ra\N$ function which
is an isomorphism) does not matter.

@@

\emph{Notational conventions.} We name elements of $\Con$ as $\Gamma$, $\Delta$,
$\Theta$, elements of $\Sub\,\Gamma\,\Delta$ as $\sigma$, $\delta$, $\nu$,
elements of $\Ty\,\Gamma$ as $A$, $B$, $C$, and elements of $\Tm\,\Gamma\,A$ as
$t$, $u$, $v$. CwF components by default support de Bruijn indices, which are
not easily readable. We use instead a nameful notation for binders in context
extension, $\Pii$ and $\lami$, e.g.\ as $(\emptycon\,\ext (a : \U) \ext (t :
\El\,a))$. We also define a type-theoretic flavor of $\appi$ for convenience:
\begin{alignat*}{3}
  & \blank\appitt\blank && :
      \Tm\,\Gamma\,(\Pii\,a\,B)\ra
      (u : \Tm\,\Gamma\,(\El\,a)) \ra \Tm\,\Gamma\,(B[\id, u])\\
  & t\appitt u && := (\appi\,t)[\id, u]
\end{alignat*}
We abbreviate non-dependent inductive $\Pii$ as $\blank\arri\blank$, and
likewise we use $\blank\arre\blank$ and $\blank\arrinf\blank$ for non-dependent
external and infinitary functions.

@@

With this, we are able to specify QIITs, and we can also interpret each
signature in an arbitrary ToS model, by applying a signature to a model.
$\Sig_j$ can be viewed as a precursor to a Church-encoding for the theory of
signatures, but we only need contexts encoded in this way, and not other ToS
components. In functional programming, this representation is sometimes called
``finally tagless'' \cite{carette2007finally}, and it is used for defining and
interpreting embedded languages.
\end{definition}

@@

We make a similar choice in the interpretation of signatures themselves: we use
structured CwFs instead of lex categories. The reason here is that CwFs allow us
to compute induction principles in strictly the same way as one would write in
type theory, since we have $\Ty$ and $\Tm$ for a primitive notion of dependent
objects and morphisms. In contrast, dependent objects in lex categories is a
derived notion, and the induction principles we get are only up to
isomorphism. This issue is perhaps not relevant from a purely categorical
perspective, but we are concerned with eventually implementing QIITs in proof
assistants, so we prefer if our semantics computes strictly.

@@

In contrast to \cite{kaposi2019constructing}, it is not sufficient to model
types as displayed flCwFs. In ibid.\ substitutions are modeled as strict
morphisms, which we cannot do because our infinitary function types and identity
types cannot be interpreted strictly, as we will see in Section
\ref{sec:infinitaryfunction}. Strict preservation is expressed with metatheoretic equality, which
is respected by all cETT constructions, but isomorphisms are not necessarily
respected. Hence, we need to add additional structure to displayed flCwFs which
expresses preservation of base isomorphisms.

@@

Moreover, $\id_{\boldsymbol{A[\sigma]}} := \id_{\bA}$,
$\sigma\circ_{\boldsymbol{A[\sigma]}}\delta := \sigma\circ_{\bA}\delta$, and
likewise components for substitution are given by corresponding components in
$\bA$.

@@

In the following, We will need three model fragments of $\bM$, which can be used
to compute notions of algebras, displayed algebras and sections respectively for
each syntactic signature. This is a rephrasing of the $\blank^{\A}$,
$\blank^{\D}$ and $\blank^{\S}$ interpretations in
\cite{kaposi2019constructing}.

@@

The basic idea is that initial algebras can be built from the terms of $\syn_j$.
For example, consider the syntactic signature for natural numbers:
\[
\NatSig := \emptycon\ext(N : \U)\ext(zero : \El\,N)\ext (suc : N
\arri \El\,N)
\]
The type $\Tm_{\syn}\,\NatSig\,(\El_{\syn}\,N)$ is isomorphic to the usual type
of natural numbers, since, intuitively, such terms can only be built from
iterated usage of $zero$ and $suc$. We build a term algebra for each
signature in this manner.

@@

So far, the underlying sets in $\bT_{\ulOmega}$ are similar to what was given in
\cite{kaposi2019constructing} in the construction of term algebras, but there is
an important difference: in ibid.\ strict equalities are used instead of
isomorphisms. In our case, isomorphisms are necessary once again because of
infinitary functions types and our identity type; we shall see this shortly. The
universe is interpreted as follows:
\begin{alignat*}{3}
  & \U_{\bT_{\ulOmega}} : (\ulnu : \Sub\,\ulOmega\,\ulGamma)(\ult : \Tm\,\ulOmega\,\U)\ra
              \Set_{j+1}\\
  & \U_{\bT_{\ulOmega}}\,\ulnu\,\ult := \Tm\,\ulOmega\,(\El\,\ult)\\
  & \El_{\bT_{\ulOmega}}\,a : (\ulnu : \Sub\,\ulOmega\,\ulGamma)
             (\ult : \Tm\,\ulOmega\,(\El\,(\ula[\ulnu])))\ra
              \llb\ula\rrb_{\bA}\,(\Gamma\,\ulnu)\\
  & \El_{\bT_{\ulOmega}}\,a\,\ulnu\,\ult := (a\,\ulnu)\,\ult
\end{alignat*}
Hence, a syntactic $\ult : \Tm\,\ulOmega\,\U$ is interpreted as a set of terms
with type $\El\,\ult$. In the interpretation of $\El$, note that
\[
  a\,\ulnu : \U_{\bT_{\ulOmega}}\,\ulnu\,(\ula[\ulnu]) \simeq_{\id}
                 \llb\ula\rrb_{\bA}\,(\Gamma\,\ulnu)
\]
hence
\[
a\,\ulnu : \Tm\,\ulOmega\,(\El\,(\ula[\ulnu])) \simeq_{\id}
           \llb\ula\rrb_{\bA}\,(\Gamma\,\ulnu)
\]
The $\simeq_{\id}$ above is just an isomorphism of sets, since it lives in
$\llb\U\rrb_{\bM}$ which was given as the flCwF of sets in Section
\ref{sec:universe}. This above isomorphism is a good summary of the
construction: the interpretation of a $\ula : \Tm\,\ulOmega\,\U$ in the term
algebra is isomorphic to a set of terms.
